### 2022-05-30 paper.tex
add Leonies part, unchanged for now

CpG dinocleotides are underrepresented in eukariotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deanimation), and the majority of existing CpG sites are methylated. Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription. Methylating DNA during differentiation is therefore conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage. One important application of single-cell bisulfite sequencing (sc-BS-Seq) therefore lies in the study of differentiation processes.

In the present paper, we discuss strategies to analyse sc-BS-Seq data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.

@@

As an example, we will us in this paper data from XX et al., who have performed sc-BS on [...] and analysed their data in the manner just described. Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100 kb. As one can see, the approach is suitable to see clear structure in the data, observe ...

@@

\begin{figure}
	A\\
	\includegraphics[width=\columnwidth]{need_for_relative_meth.png}
	B\\
	.\hspace{.39cm}\includegraphics[width=.83\columnwidth]{smoothing_residuals.png}
\caption{\small \textbf{Effect of read position:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified. Two cells cover differing parts of the interval with one read each. If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells. \textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained. To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
\label{smoothres}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{vmr.png}
	\caption{\small \textbf{Finding variably methylated regions (VMR):} The chromosomes are divided up into overlapping windows (first four shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.\ \ref{smoothres}B. Then, the variance of these values are calculated (each point represents one of the overlapping windows), a threshold (red line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold. Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions''.}
	\label{vmr}
\end{figure}

@@

We have implemented the approach just described in a Python package, called ``scbs'', that offers the following functionality: ...

@@

Now, we can compare the simple analysis described initially with an analysis using our method, in order to demonstrate that it improves results.

>>>>

CpG dinocleotides are underrepresented in eukariotic genomes (due to their vulnerability to mutation by conversion to TpG due to accidental deanimation), and the majority of existing CpG sites are methylated. Presence or absence of the methyl group at CpG sites is believed to have effect on gene regulation, with methylation inhibiting transcription. Methylating DNA during differentiation is therefore conjectured to be an important mechanism to silence genome regions that are not needed in the target lineage. One important application of single-cell bisulfite sequencing (scBS) therefore lies in the study of differentiation processes.

In the present paper, we discuss strategies to analyse scBS data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.

@@

As an example, we will us in this paper data from XX et al., who have performed scBS on [...] and analysed their data in the manner just described. Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100 kb. As one can see, the approach is suitable to see clear structure in the data, observe ...

@@

\begin{figure}
	A\\ 
	\includegraphics[width=\columnwidth]{figures/Fig_residuals_A.png}
	B\\
	\includegraphics[width=\columnwidth]{figures/Fig_residuals_B.png}
\caption{\small \textbf{Effect of read position:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified. Two cells cover differing parts of the interval with one read each. If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells. \textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained. To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
\label{smoothres}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
	\caption{\small \textbf{Finding variably methylated regions (VMR):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.\ \ref{smoothres}B. Then, the variance of these values are calculated (each point represents one of the overlapping windows), a threshold (red line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold. Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions''.}
	\label{vmr}
\end{figure}

@@

We have implemented the approach just described in a Python package, called ``scbs'', that offers the following functionality:

\subsection{Efficient storage and access of whole-genome scBS data}

\subsection{Identification and filtering of low-quality cells}

\subsection{Quantification of methylation in genomic intervals}

@@

To demonstrate an improvement of results, we compare previously published analyses described above with our analysis method.
First, we use a dataset by \citealp{Argelaguet_2019}. In their work, they generated a single cell methylome dataset of mouse embryo cells across different stages of development (E4.5 – E7.5). They calculated the correlation between RNA expression intensity and corresponding DNA methylation at the promoter region to elucidate the implications of the epigenome on RNA expression during development. As promoters, they used DNA regions of $\pm$ 2kb around transcriptions start sites (TSS).

--The following paragraph potentially needs to be changed when new plots are made--

In our analysis, we compare correlations between RNA expression intensity and DNA methylation of these promotor regions to correlations between RNA expression intensity and DNA methylation of the closest variable methylation regions that we identified with scbs. In figure \ref{figure:correlation}, we show that correlation of RNA expression intensity to gene methylation around the TSS $\pm$ 2kb is less strong compared to the correlation at the closest variable region. This effect can be observed with mehtylation fractions (figure \ref{figure:correlation}A) or shrunken residuals (figure \ref{figure:correlation}B) as data inputs.

\begin{figure}[!htp]
	A\\
	\hspace{.2cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
	B\\
	\hspace{.2cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
	C\\
	\hspace{.39cm}\includegraphics[width=\columnwidth]{part_leonie_git/leonie_plots/comparison_meth_RNA.pdf}

	\caption{\small \textbf{Correlation of DNA methylation and RNA expression intensity}. All plots are based on the data by \citealp{Argelaguet_2019}.
	\textbf{(A)} Correlation of each gene's RNA expression intensity to methylation fraction of TSS $\pm$ 2kb (y-axis) or to methylation fraction of closest variable gene.
	\textbf{(B)} Correlation of each gene's RNA expression intensity to shrunken residual of methylation of TSS $\pm$ 2kb (y-axis) or to shrunken residual of closest variable gene.
	\textbf{(C)} Mean gene methylation of Ube2l6 gene (ENSMUSG00000027078, marked in yellow in A \& B). Cells were into 4 groups based on their RNA expression (group 0: cells with no expression of gene, group 1-3: cells that express gene divided into three equally large groups with group 3 having the highest expression). Mean methylation level of each group was smoothed with a tricube kernel of bandwidth = 2000.}
	\label{figure:correlation}
\end{figure}

As an example to illustrate this effect more clearly, we picked a gene with significantly stronger correlation at the closest variable region than at the TSS. This gene is marked on figure \ref{figure:correlation}A and B with a yellow dot. We divided all cells into four groups of cells with no, low, medium and high expression of this exemplary gene Ube216 and plotted a smoothed methylation level per group. As shown in figure \ref{figure:correlation}C, the promoter region is sparsely methylated in all four RNA expression groups and is therefore a poor region to make conclusions about the relation of DNA methylation and RNA expression intensity. However, a region close to the TSS, identified as the closest highly variable methylation region with scbs, shows a much clearer negative correlation between RNA expression intensity and DNA methylation.

To further demonstrate the benefit of detecting variable methylation regions, we used a second dataset, published by \citealp{Luo_2017}. In their work, \citealp{Luo_2017} generated methylomes from single cell neuronal nuclei in mice to identify genomic regulatory elements across neuron types. To compare how well cell types are separated by different analysis techniques, we processed the same dataset into 2-dimensional UMAPs several times. Subsequently, we compared the results by assigning a score to each output based on cell type separation in the final 15 principal components. We see an improvement of cell separation when comparing the analysis of variable methylation regions to 100 kb regions (\ref{figure:score}A and C). The difference of using shrunken residuals instead of methylation fractions to assess the methylation per region is not as significant. However, when emulating a sparse dataset with less cells by randomly reducing the dataset to 500 cells, it becomes clear that using shrunken residuals over methylation fractions also improves the outcome of our analysis (\ref{figure:score}B and D).

\begin{figure}[!htp]
	A\\
	\hspace{.3cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf} \\
	B\\
	\hspace{.3cm}\includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf} \\
	C\\
	\hspace{.39cm}\includegraphics[width=\columnwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}
	D\\
	\hspace{.39cm}\includegraphics[width=\columnwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
	\caption{\small \textbf{Analysis comparison}. All plots are based on the data by \citealp{Luo_2017}. \textbf{(A)} Nearest neighbor count score for complete dataset using four different analysis techniques and a variable number of regions as input. Exemplary UMAPs for encircled scores are shown in (C). \textbf{(B)} Nearest neighbor count score for the same dataset reduced to 500 cells using four different analysis techniques and a variable number of regions as input. Exemplary UMAPs for encircled scores are shown in (D). \textbf{(C)} Two exemplary UMAPs using 1000 regions of the full dataset and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation regions (right) \textbf{(D)} Two exemplary UMAPs using 1000 regions of the reduced dataset and methylation fraction of 100 kb windows (left) or shrunken residuals of variable methylation regions (right).}
	\label{figure:score}
\end{figure}
